---
title: "BibAnalysis"
author: "O.W."
date: "2025-12-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)

setwd("C:/Users/PC/OneDrive - University of the Sunshine Coast/CatPaper")

pacman::p_load(tidyverse,
               data.table,
               patchwork,
               lme4,
               lmerTest)

# plotting variables
bib_colours <- c("On" = "#e89fbf",
                 "Off" = "#0097b2")
prediction_colours <- c("Locomotion" = "#e89fbf",
                        "high" = "#d8907c",
                        "medium" = "#e6c078",
                        "low" = "lightgreen",
                        "inactive" = "#0097b2")

individual_colours <- c("plum", "#e89fbf", "rosybrown1", "#d8907c","#e6c078", "khaki2", "lightgreen","#0097b2", "skyblue4")

my_theme <- function() {
  theme_minimal(base_size = 10, base_family = "serif") +
    theme(
      panel.border = element_rect(color = "black", linewidth = 1.5, fill = NA),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.background  = element_rect(fill = "transparent", colour = NA),
      panel.background = element_rect(fill = "transparent", colour = NA),
      axis.line = element_line(color = "black"),
      axis.ticks = element_line(color = "black"),
      axis.title = element_text(size = 20),
      axis.text = element_text(size = 20) 
    )
}

# setting up data
# load in the data and clean it up
final_data <- fread("C:/Users/PC/OneDrive - University of the Sunshine Coast/CatPaper/Output/Results/Final_predictions.csv") %>%
  filter(!bib_status == "") %>%
  mutate(hour = hour(time),
         date = as.Date(time),
         prediction = ifelse(prediction == "FastLocomotion", "Locomotion", prediction)) %>%
  na.omit()
# define some factor levels
final_data$bib_status <- factor(final_data$bib_status, levels = c("Off", "On"))
final_data$prediction <- factor(final_data$prediction, levels = c("Locomotion", "high", "medium", "low", "inactive"))
```

## Summarising data
What data was used, and how much of it?
```{r}
IDs <- length(unique(final_data$ID))
duration_total <- nrow(final_data)/60/60/24 # days
duration_bibon <- nrow(final_data[final_data$bib_status == "On"])/60/60/24
duration_biboff <- nrow(final_data[final_data$bib_status == "Off"])/60/60/24

print(cat("Individuals:", IDs, "\n",
          "Total sampling hours:", nrow(final_data)/60/60, "\n",
          "Total sampling days:", duration_total, "\n", 
          "Total bib on days:", duration_bibon,  "(", round(duration_bibon/duration_total,2), ")",  "\n",
          "Total bib off days:", duration_biboff, "(", round(duration_biboff/duration_total,2), ")"))

```

## Model Performance
Before making predictions with the model, we should assess the performance from the cross-validation.
```{r}
Locomotion_CV <- fread("C:/Users/PC/OneDrive - University of the Sunshine Coast/CatPaper/Output/ModelBuilding/Locomotion_CrossValidation.csv")

Locomotion_performance <- Locomotion_CV %>% select(-fold, -Accuracy) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>%
  mutate(Behaviour = "Locomotion")
Locomotion_performance
```

## Energy Over the Day
How does mean activity change throughout the day depending on the bib status.
```{r daily_energy, fig.width=12, fig.height=5}
# wrangle the data
plot_data <- final_data %>%
  group_by(ID, date, hour, bib_status) %>%
  summarise(mean = mean(mean_VDBA))

plot_data2 <- final_data %>%
  group_by(ID, bib_status, prediction) %>%
  na.omit() %>%
  summarise(mean = mean(mean_VDBA))

# make the plots
p1 <- ggplot(plot_data, aes(y = mean, x = as.numeric(hour), colour = bib_status)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE, span = 0.3) +
  xlab("Hour") +
  ylab("Mean VDBA (g)") +
  my_theme() + 
  scale_colour_manual(values = bib_colours)

p2 <- ggplot(plot_data2, aes(x = prediction, y = mean, fill = bib_status)) +
  geom_boxplot() +
  xlab("Bib Status") +
  ylab("Mean VDBA (g)") +
  my_theme() + 
  scale_fill_manual(values = bib_colours)
  
plot1 <- (p1 + p2) +
    plot_layout(guides = "collect") &
    theme(
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.title = element_text(size = 10),
      legend.text  = element_text(size = 10),
      legend.key.size = unit(0.4, "lines"),
      legend.spacing.x = unit(0.3, "cm"),
      legend.spacing.y = unit(0.2, "cm")
    )
  
plot1

# do the stats
stat_data <- final_data %>%
  group_by(ID, date, hour, bib_status, prediction) %>%
  summarise(mean = mean(mean_VDBA))
m <- lmer(
  log(mean) ~ bib_status * prediction + 
    (1 | ID),
  data = stat_data
)
anova(m)
summary(m)

```

## Behavioural Frequency
Does the frequency of behaviours change throughout the day?
```{r proportion_behaviours}
plot_data3 <- final_data %>%
  group_by(hour, bib_status, prediction) %>%
  summarise(total_count = n()) %>%
  mutate(prop = total_count / sum(total_count))

plot_data4 <- final_data %>%
  group_by(bib_status, prediction) %>%
  summarise(total_count = n()) %>%
  group_by(bib_status) %>%
  mutate(prop = total_count / sum(total_count))


p3 <- ggplot(plot_data3, aes(x = hour, y = prop, fill = prediction)) +
  geom_area(position = "stack", alpha = 0.8) +
  facet_wrap(~ bib_status, ncol = 1, strip.position = "bottom") +  # move facet titles
  scale_fill_manual(values = prediction_colours) +
  labs(x = "Hour", y = "Proportion of behaviours") +
  my_theme()

p4 <- ggplot(plot_data4, aes(x = bib_status, y = total_count, fill = prediction)) +
  geom_bar(stat = "identity", position = "fill", alpha = 0.8) +
  scale_fill_manual(values = prediction_colours) +
  labs(y = NULL, x = "Bib Status") +
  my_theme()


plot2 <- (p3 + p4) +
    plot_layout(guides = "collect") &
    theme(
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.title = element_text(size = 10),
      legend.text  = element_text(size = 10),
      legend.key.size = unit(0.4, "lines"),
      legend.spacing.x = unit(0.3, "cm"),
      legend.spacing.y = unit(0.2, "cm")
    )
  
plot2

# stats
stat_data2 <- final_data %>%
  group_by(ID, bib_status, prediction) %>%
  summarise(total_count = n()) %>%
  mutate(prop = total_count / sum(total_count)) %>%
  group_by(ID, bib_status) %>%
  mutate(prop = total_count / sum(total_count))

glmer(
  prediction ~ bib_status + (1 | ID),
  data = stat_data2,
  family = binomial
)

```

## Inter-Individual Variation
Did the bib affect cats differently? and did it depend on their morphology or personality?

```{r fig.width=18, fig.height=6}
# plot of the slopes
stat_data3 <- final_data %>%
  group_by(ID, bib_status, hour, date) %>%
  summarise(mean = mean(mean_VDBA), )
slopes <- ggplot(stat_data3, aes(bib_status, mean, colour = ID, group = ID)) +
  stat_summary(fun = mean, geom = "line", linewidth = 2) +
  stat_summary(fun = mean, geom = "point", size = 2) +
  my_theme() +
  labs(x = "Bib status", y = "Mean VDBA") +
  scale_colour_manual(values = individual_colours) +
  theme(legend.position = "none")

leg_lengths <- fread("C:/Users/PC/OneDrive - University of the Sunshine Coast/CatPaper/Data/CatInfo.csv")
stat_data3 <- merge(stat_data3, leg_lengths, by = "ID")

# plot of the sploes against leg length
ind_means <- stat_data3 %>%
  group_by(ID, bib_status) %>%
  summarise(mean_pred = mean(mean), .groups = "drop")
ind_slopes <- ind_means %>%
  pivot_wider(names_from = bib_status, values_from = mean_pred) %>%
  mutate(slope = On - Off)  # difference from Off to On
# Add FrontLegLength (assuming it's constant per individual)
frontleg_df <- stat_data3 %>%
  select(ID, FrontLegLength, BodyLength) %>%
  distinct()
ind_slopes <- left_join(ind_slopes, frontleg_df, by = "ID")

leg_length <- ggplot(ind_slopes, aes(x = FrontLegLength, y = slope, colour = ID)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "#0097b2", linewidth = 2) +
  labs(x = "Front Leg Length (cm)", y = "Change in mean prediction (On − Off)") +
  my_theme() +
  scale_colour_manual(values = individual_colours) + theme(legend.position = "none")

body_length <- ggplot(ind_slopes, aes(x = BodyLength, y = slope, colour = ID)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "#0097b2", linewidth = 2) +
  labs(x = "Body Length (cm)", y = "Change in mean prediction (On − Off)") +
  my_theme() +
  scale_colour_manual(values = individual_colours) + theme(legend.position = "none")

plot3 <- (slopes + leg_length + body_length) +
    plot_layout(guides = "collect") &
    theme(
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.title = element_text(size = 10),
      legend.text  = element_text(size = 10),
      legend.key.size = unit(0.4, "lines"),
      legend.spacing.x = unit(0.3, "cm"),
      legend.spacing.y = unit(0.2, "cm")
    )
  
plot3


# individual effects
m <- lmer(log(mean) ~ bib_status + (1 | ID), data = stat_data3)
m_rs <- lmer(log(mean) ~ bib_status + (bib_status | ID), data = stat_data3)
anova(m, m_rs)

# effect of morphology
mod_interaction <- lmer(mean ~ bib_status * FrontLegLength + 
                         bib_status * BodyLength + 
                         (bib_status | ID),
                         data = stat_data3)
summary(mod_interaction)

```

## Assessing learning curve 
Was the effect of the bib strongest on day 1 of wearing the bib
```{r}
setDT(final_data)

# Only bib On data
bib_data <- final_data[bib_status == "On"]

# Day relative to first bib-on day
bib_data[, first_bib_day := min(date), by = ID]
bib_data[, day := as.integer(date - first_bib_day) + 1]

# Daily summary for bib On
daily_summary <- bib_data[, .(
  mean_VDBA = mean(mean_VDBA),
  se_pred = sd(mean_VDBA)/sqrt(.N)
), by = .(ID, day)]

# Bib Off baseline
bib_off <- final_data %>%
  filter(bib_status == "Off") %>%
  group_by(ID) %>%
  summarise(mean_VDBA = mean(mean_VDBA), .groups = "drop") %>%
  mutate(day = 0, bib_status = "Off")

# Combine
combined <- bind_rows(
  bib_off %>% rename(bib_status = bib_status),
  daily_summary %>% mutate(bib_status = "On")
)

# Plot
ggplot(combined, aes(x = day, y = mean_VDBA, colour = ID)) +
  geom_point(size = 2) +
  geom_line(aes(group = ID), alpha = 0.3, linewidth = 1) +
  stat_summary(aes(group = bib_status), fun = mean, geom = "line", size = 1.5) +
  scale_colour_manual(values = individual_colours) +
  labs(x = "Day since bib worn", y = "Mean VDBA", colour = "ID") +
  my_theme()

# stats to look at the efefct of day
mod_day <- lmer(mean_VDBA ~ bib_status * factor(day) + (1 | ID),
                data = combined)

summary(mod_day)

```



